Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-04-27 concurrency/CONCUR 2023/paper.tex
section 3 edits

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}

@@

Recall that in the lock-coupling pattern, each thread consistently retains at least one lock while traversing between nodes, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, the thread must acquire a lock for the current node $\texttt{pn->n}$ and obtain the resource safeguarded by the lock. This resource, represented as $\texttt{R(pn->n)}$, is parameterized by a node pointer $\texttt{pn->n}$ and describes the node's concrete information.

As mentioned earlier, we can define a piece of ghost state, $\texttt{ghost\_node t}$, connecting the node in memory to the public abstract state of the data structure. Specifically, we can have one half of the ghost state contained in the resource $\texttt{R(pn->n)}$, and the other half in the abstract state $\texttt{Node t}$. When proving the atomic triple above, at the linearization point where we access $\texttt{Node}$, we combine both halves of the ghost state, update them to reflect the new state of the concrete data structure, and then fulfill the atomic postcondition with one half of the ghost state, returning the other half to the lock invariant.

In the specification, the local pre-condition includes both the representation of the node $\nodeboxrep(\texttt{pn->n})$ and the resource $\texttt{R(pn->n)}$. When the \texttt{traverse} function operates on the entire search structure, it generates a subsequent node, \texttt{n'}, such that the key \texttt{k} falls within the range of node \texttt{n'}. The local post-condition is expected to encompass the representation of the new node $\nodeboxrep(\texttt{pn->n'})$, the resource $\texttt{R(pn->n)}$, and a Boolean variable, \texttt{res}, which indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not. The key to verifying the \texttt{traverse} function is identifying the loop invariant for the top-level loop. We define the loop invariant for the lock-coupling style \texttt{traverse} function as follows:
\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range)}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
\end{mathpar}
The loop invariant mentioned above include specific existential variables (such as $\texttt{n'}$) that define the state of each iteration throughout the loop. The \lstinline{traverse} function stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 6 in Figure \ref{traverse_lock_b}). At this point, the specification indicates that we have $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$, enabling us to prove the post-condition. On the other hand, the function terminates if \lstinline{findNext} is unable to locate a subsequent node with the key value  $\texttt{k}$ within its set of key values (line 11 in Figure \ref{traverse_lock_b}). In this case, the specification clarifies that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} \neq \texttt{NULL}$ which also enables us to prove the post-condition.

The proof begins with $\nodeboxrep \texttt{(pn->n)}$, which comprises a lock invariant $\texttt{pn->n->lock} \lockvar \texttt{R(pn->n)}$, and a resource $\texttt{R(pn->n)}$ acquired before invoking the \lstinline{traverse} function. After \lstinline{findNext} successfully detects the next node (for example, \texttt{n'}) to visit, we acquire the lock of node \texttt{n'} (line 14 in Figure \ref{traverse_lock_b}). It means we can get the resource $\texttt{R(pn->n')}$ from $\texttt{pn->n'->lock} \lockvar \texttt{R(pn->n')}$. Subsequently, we release the lock of the current node \texttt{p}. It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}). By acquiring the lock for the next node \texttt{n'} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it (hand-over-hand locking). 
The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

\than{I feel somewhat perplexed while attempting to articulate the situation mentioned above. Can someone help me fix or explain it more?}

\subsubsection{Give-up Template}
\label{traverse_proof_giveup}
Moving forward with the proofs, we define the following specification for the traverse function within the give-up template, as depicted in Figure \ref{fig:traverse_giveup}.

@@

In contrast to the more complex specification of the lock-coupling style \texttt{traverse} function, the give-up template's \texttt{traverse} function has a comparatively simpler specification. The local pre-condition for the specification solely comprises $\infp(\texttt{pn->n})$, which is a footprint predicate signifying that node \texttt{n} resides within the search structure's footprint. Subsequently, the \texttt{traverse} function operates on the entire search structure, returning a subsequent node \texttt{n'} where \texttt{k} lies within the node \texttt{n'}'s range. Therefore, the local post-condition is expected to assert that \texttt{n'} is also positioned within the footprint predicate, that the resource \texttt{R(pn->n')} embodies the concrete information concerning node \texttt{n'}, and that a Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.

In contrast to the lock-coupling style, where each lock is associated with a lock-invariant—an arbitrary separation logic predicate representing the protected resource acquired by a thread when the lock is obtained and must be restored upon lock release—the give-up template has a different approach. In this method, every node possesses a lock bit at location $\ell$, which is set to \texttt{true} if and only if the node is locked. This protocol is encoded into the definition of the abstract state $\treerep$. The protocol is formally defined as $\texttt{inv\_for\_lock} \ \ell \ \texttt{R} \triangleq \exists b. \ \ell \mapsto b \ \ast \ \texttt{if } b \ \texttt{then emp else R}$, where a lock at location $\ell$ safeguards resources denoted by \texttt{R}. When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$. On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.

Before proving the \texttt{traverse} function, let's define the loop invariant for the given template of the \texttt{traverse} function as follows:
\begin{mathpar} \texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{mathpar}
To begin the proof, we start with the $\infp (\texttt{pn->n})$ predicate from the local pre-condition. After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$ and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

If \lstinline{findNext} cannot find a subsequent node with the key value $\texttt{k}$ within its set of key values, the function terminates. Otherwise, it successfully detects the next node (for example, \texttt{n'}) to visit, giving a predicate that asserts \texttt{n'} is in the footprint of the structure, $\infp(\texttt{pn->n})$. After releasing the lock of the current node \texttt{p} (since \texttt{p} and \texttt{n} point to the same memory address - line 6 in Figure \ref{traverse_giveup_b}), we can prove that it satisfies the loop invariant $\texttt{traverse\_inv(pn)}$ and complete this branch of the proof. In the \texttt{else} branch of the call to \texttt{inRange}, since \texttt{k} is not in the range of the current node \texttt{p}, the function gives up and relinquishes the lock on \texttt{p}, then goes back to the root of the data structure to retry.

>>>>

\subsubsection{Lock-coupling Template}
\label{traverse_proof_lock}
\wm{Does this organization work, or would it be better to prove each template as we introduce it?}

@@

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{pn->n}$ and obtain the resources $\texttt{R(pn->n)}$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $R$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. \wm{Should we replace this with a concrete definition of the invariant?}

%As mentioned earlier, we can define a piece of ghost state, $\texttt{ghost\_node t}$, connecting the node in memory to the public abstract state of the data structure. Specifically, we can have one half of the ghost state contained in the resource $\texttt{R(pn->n)}$, and the other half in the abstract state $\texttt{Node t}$. When proving the atomic triple above, at the linearization point where we access $\texttt{Node}$, we combine both halves of the ghost state, update them to reflect the new state of the concrete data structure, and then fulfill the atomic postcondition with one half of the ghost state, returning the other half to the lock invariant.

In the specification, the local precondition includes both the node handle $\nodeboxrep(\texttt{pn->n})$ and the contents $\texttt{R(pn->n)}$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{pn->n'})$ \wm{This looks weird: \lstinline{pn} doesn't have an \lstinline{n'} field}, its contents $\texttt{R(pn->n')}$, and a Boolean variable \texttt{res} indicating whether \texttt{pn->n'->t} is \texttt{NULL} or not \wm{Indicate the relationship between \lstinline{res} and the \lstinline{flag} return value}. The key to the correctness of the \texttt{traverse} function the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{mathpar}\texttt{traverse\_inv(pn)} \triangleq\ \exists \ \texttt{n'}.\ \texttt{(k} \in \texttt{range)}\ \land \ \nodeboxrep\texttt{(pn->n')}  \ \ast \  \texttt{R(pn->n')}   
\end{mathpar}\wm{Where does \lstinline{range} come from here?}

Given this invariant, the proof proceeds as follows. \wm{Should we consider giving annotated code after all?} We begin by checking whether the current node is null (line 5 of Figure~\ref{traverse_lock_b}); if it is, we have found the leaf where \lstinline{k} belongs, and can prove the postcondition with \lstinline{res = true}. Otherwise, we use \lstinline{findNext} to find a new node \lstinline{n'} to visit (line 8), which is stored in \lstinline{pn->n}. If \lstinline{findNext} returns 0, we have found a node containing \lstinline{k}, and can prove the postcondition with \lstinline{res = false}. Otherwise, we acquire \lstinline{pn->n->lock}, gaining access to the resources $\texttt{R(pn->n)}$, and then release the lock of the current node \texttt{pn->p} and return its resources. %It is important to note that \texttt{p} point to the same memory address as \texttt{n} (line 4 in Figure \ref{traverse_lock_b}).
By acquiring the lock for the next node \texttt{n} before releasing the lock for the current node \texttt{p}, we ensure that the connection between the two nodes remains intact and unaltered while we traverse it, and re-establish \lstinline{traverse_inv} for the new values in \lstinline{pn}. 
%The proof concludes by verifying that the \lstinline{traverse} function fulfills the post-condition of the specification mentioned earlier.

\subsubsection{Give-up Template}
\label{traverse_proof_giveup}

The atomic specification for the give-up template's \lstinline{traverse} is shown in Figure \ref{fig:traverse_giveup}.

@@

The give-up template's \texttt{traverse} specification is simpler than that of the lock-coupling template. The local precondition include any locks, and consists only of $\infp(\texttt{pn->n})$, a footprint predicate asserting that node \texttt{n} is included in the search structure's abstract state $m$. As before, the \texttt{traverse} function will navigate the structure and return a node \texttt{n'} whose range includes \texttt{k}, as well as acquiring \lstinline{n'}'s lock and returning its contents \texttt{R(pn->n')}. Also as before, the Boolean \texttt{res} indicates whether \texttt{pn->n'->t} is \texttt{NULL} or not.

\wm{We really need to explain why we have this difference, or get rid of it.} In contrast to the lock-coupling style, where each lock is associated with a lock-invariant, in the give-up template we treat locks more abstractly. For each node, we track whether its lock is held or not, and the abstract state $\treerep$ holds the contents of all nodes whose locks are not currently held. Formally, we write \[\texttt{inv\_for\_lock} \ \ell \ \texttt{R} \triangleq \exists b. \ \ell \mapsto b \ \ast \ \texttt{if } b \ \texttt{then emp else R}\]\wm{Consider using mathsf for logic and reserving texttt for C code.}
%where a lock at location $\ell$ safeguards resources denoted by \texttt{R}. When the lock is held, meaning the node is locked ($\ell \mapsto \texttt{true}$), threads can remove resources \texttt{R} from $\treerep$. On the other hand, when the lock is not held ($\ell \mapsto \texttt{false}$), resources \texttt{R} are required to be part of the invariant.
and $\treerep$ contains an \texttt{inv\_for\_lock} assertion for each node in the data structure. As before, $R$ contains the concrete contents of each node, but also includes the \lstinline{min} and \lstinline{max} fields that we use to implement the \lstinline{inRange} check.

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \texttt{traverse\_inv(pn)} \triangleq \ \exists \ \texttt{n'}.\ \infp (\texttt{pn->n}) \ast \infp (\texttt{pn->n'})   
\end{mathpar}
With this invariant in hand, the proof of \lstinline{traverse} proceeds as follows. We start with the $\infp (\texttt{pn->n})$ predicate from the local precondition. After acquiring the lock of the node $\texttt{pn->n}$ (line 5 in Figure \ref{traverse_giveup_b}), we obtain the resource $\texttt{R(pn->n)}$, which allows us to perform the \lstinline{inRange} check on line 7. If the check fails (lines 20-23), we immediately release the lock and start over from the beginning. Otherwise, we proceed as before: if the current node's contents are \texttt{NULL}, we have found the empty leaf where \lstinline{k} belongs (and can satisfy the postcondition); if \lstinline{findNext} returns 0, we have found the node containing \lstinline{k} (and can satisfy the postcondition); otherwise, we release our lock and re-establish \lstinline{traverse_inv} for the new node indicated by \lstinline{findNext}. In this case, we do not hold the lock on the new node, but only the $\infp$ assertion indicating that the new node is in the data structure. % and add the additional predicate $\texttt{k} \in \texttt{range(n)}$ in the \texttt{then} of \texttt{inRange}. Similar to the proof of the lock-coupling template's \texttt{traverse}, the traversal stops when it reaches the base case, which occurs when the current node is \lstinline{NULL} (line 8 in Figure \ref{traverse_giveup_b}). At this point, we can prove the post-condition using the specification, which indicates that $\texttt{k} \in \texttt{range(n')} \ast \texttt{pn->n'->t} = \texttt{NULL}$.

%If \lstinline{findNext} cannot find a subsequent node with the key value $\texttt{k}$ within its set of key values, the function terminates. Otherwise, it successfully detects the next node (for example, \texttt{n'}) to visit, giving a predicate that asserts \texttt{n'} is in the footprint of the structure, $\infp(\texttt{pn->n})$. After releasing the lock of the current node \texttt{p} (since \texttt{p} and \texttt{n} point to the same memory address - line 6 in Figure \ref{traverse_giveup_b}), we can prove that it satisfies the loop invariant $\texttt{traverse\_inv(pn)}$ and complete this branch of the proof. In the \texttt{else} branch of the call to \texttt{inRange}, since \texttt{k} is not in the range of the current node \texttt{p}, the function gives up and relinquishes the lock on \texttt{p}, then goes back to the root of the data structure to retry.
